A Meta-Analysis of Music Emotion Recognition Studies

### 2025-04-02 manuscript/manuscript.qmd
added revisions to conclusions

Emotional engagement is a key reason why people engage with music in their everyday activities, and it is also why music is increasingly being used in various health applications [@juslin2022emotions; @agres2021music]. In recent years, significant advances have been made in music information retrieval, particularly in music emotion recognition (MER) tasks [@gomez2021; @panda2020audio]. Improvements to modelling techniques, datasets, and available features have created new opportunities to improve the accuracy and reliability of MER systems developed to predict emotion labels or ratings in music using audio features. Over the past 25 years, these studies have established the types of emotions that listeners perceive and recognize in music. In the last 15 years, research has increasingly focused on tracing these recognized emotions back to specific musical components, such as expressive features [@lindstrom2003expressivity], structural aspects of music [@eerola_friberg_bresin_2013; @anderson2022ex; @grimaud_eerola_2022], acoustic features [@yang2008; @panda2013multi; @saari_et_al_2015; @eerola2011c; @panda2020audio], or emergent properties revealed through deep learning techniques [@er2019music; @sarkar2020recognition].

Despite increased interest in MER studies, there is no consensus on the extent to which emotions can accurately be recognized by computational models. The current literature presents a diverse and mixed picture regarding the success of models in predicting emotions within the affective circumplex -- valence and arousal-- [@Russell1980] and in classifying distinct emotion categories [@fu2010survey].

@@

@@

<!-- first para is your text from ICMPC paper -->

Research on Music Emotion Recognition has steadily increased in popularity since the early 2000s, with technological advancements facilitating sharing of data sets, analysis tools, and music stimuli. Public forums like the Music Information Retrieval Exchange (MIREX) have facilitated collaborations between computer scientists, musicologists, and psychologists alike—spurring improvements in performance. Despite increasing complexity in the types of models and data sets used, no extant study has evaluated the relative merits of the tools, stimuli, and feature sets employed in MER studies. The present study offers the first meta-analysis of music emotion recognition.

@@

Second, there is no single modelling technique that seems to arise on top, although linear models and random forests perform best in regression studies, and neural networks and support vector machines excel in classification tasks. We note that the model accuracy is surprisingly little affected by the number of features or the dataset size—likely the result of cross-validation techniques used to avoid overfitting—but the heterogeneity of the materials used in different studies may also mask substantial differences. For instance, the smaller datasets tend to use linear models and deliver higher model accuracy than larger datasets and those with a large number of features. We surmise that this might relate to disciplinary differences. For example, the smaller datasets often come from music psychology studies, which put a premium on data quality (quality control of the ground-truth data and features) rather than on data size and model techniques. Successful quality control is thought to decrease random errors inherent in human judgments. This argument is largely consistent with the analysis of the studies divided across journal types (psychology vs. engineering).

Third, there is little work on the relevance of the features and how these impact the accuracy of the models. Unfortunately, in many cases, it is impossible to attribute the sources of error to features (as opposed to target emotions obtained from participants or from the modelling architecture, cross-validation, or the size of data), as the studies so far have not compared the feature sets systematically, nor has a comparison of the datasets with identical features has been carried out. In the future, it would be advantageous to systematically assess these sources of error in MER studies to allow us to focus on where and how significant improvements can be made.

@@

Future reports should contain viable information about the data, models, and success metrics. The modelling process should include a description of cross-validation, feature extraction technique, feature reduction (if used), and actual accuracy measures. We would recommend future studies to use Matthews correlation coefficient ($MCC$) for classification studies as currently there are numerous metrics (*F~1~*, precision, recall, balanced accuracy, etc.), which do not operate in the same range. The situation for regression studies tends to be less volatile, but model accuracy may be reported with error measures (Mean Squared Error, Root Mean Squared Error, or Mean Absolute Error, or $MSE$, $RMSE$, $MAE$), which cannot be converted to relative measure such as $R^2$ without knowing the variance of the data.

It would be also highly beneficial to share models and features transparently. If a study relies on published datasets, as most do, sharing the actual model (as code/notebooks) and features would allow for a more direct comparison of model techniques for the same dataset. Also, not all datasets are shared [@akiki2021; @santana2020; @hizlisoy2021mu; @zhang2017fe], or at least the sharing is only done partially. Standardizing reporting of datasets to include subsections explaining detailed information about stimuli (including genre, duration, sampling rate, and encoding format), features (types, extraction/analysis software, quantity of features, transformations, reduction methods), and models (types, tuning parameters, cross-validation) will enable more accurate comparisons between studies. The reporting in @grekow2018au serves as a useful example by providing enough detail in these areas to facilitate reproducibility. Although copyright restrictions may limit the sharing of some datasets, features and examples should be made available.

>>>>

Emotional engagement is a key reason why people engage with music in their everyday activities, and it is also why music is increasingly being used in various health applications [@juslin2022emotions; @agres2021music]. In recent years, significant advances have been made in music information retrieval (MIR), particularly in music emotion recognition (MER) tasks [@gomez2021; @panda2020audio]. Music Emotion Recognition (MER) is an interdisciplinary field that combines computer science, psychology, and musicology to identify the emotions conveyed by music. Research in this area involves developing computational models capable of recognizing emotions from musical content. The emotional attributions of music are based on various theoretical frameworks for emotion and require annotated datasets to build and train these models. Improvements to modelling techniques, datasets, and available features have created new opportunities to improve the accuracy and reliability of MER systems developed to predict emotion labels or ratings in music using audio features. Over the past 25 years, these studies have established the types of emotions that listeners perceive and recognize in music. In the last 15 years, research has increasingly focused on tracing these recognized emotions back to specific musical components, such as expressive features [@lindstrom2003expressivity], structural aspects of music [@eerola_friberg_bresin_2013; @anderson2022ex; @grimaud_eerola_2022], acoustic features [@yang2008; @panda2013multi; @saari_et_al_2015; @eerola2011c; @panda2020audio], or emergent properties revealed through deep learning techniques [@er2019music; @sarkar2020recognition].

<!-- added a MER definition to 2nd sentence above -->

Despite increased interest in MER studies, there is no consensus on the extent to which emotions can accurately be recognized by computational models. The current literature presents a diverse and mixed picture regarding the success of models in predicting emotions within the affective circumplex -- valence and arousal-- [@yang2011] and in classifying distinct emotion categories [@fu2010survey].

@@

<!-- DISCUSS the interpretation above -->

@@

Research on Music Emotion Recognition has steadily increased in popularity since the early 2000s, with technological advancements facilitating sharing of data sets, analysis tools, and music stimuli. Public forums like the Music Information Retrieval Exchange (MIREX) have facilitated collaborations between computer scientists, musicologists, and psychologists alike -- spurring improvements in performance. Despite the increasing complexity of models and datasets, no existing study has rigorously compared the overall success of Music Emotion Recognition research using standardized metrics, nor has there been an analysis of the relative merits of the model techniques and feature sets employed. This study presents the first meta-analysis of music emotion recognition.

@@

Second, there is no single modelling technique that seems to arise on top, although linear models and random forests perform best in regression studies, and neural networks and support vector machines excel in classification tasks. We note that the model accuracy is surprisingly little affected by the number of features or the dataset size—likely the result of cross-validation techniques used to avoid overfitting—but the heterogeneity of the materials used in different studies may also mask substantial differences. For instance, the smaller datasets tend to use linear models and deliver higher model accuracy than larger datasets and those with a large number of features. We surmise that this might relate to disciplinary differences. For example, the smaller datasets often come from music psychology studies, which put a premium on data quality (quality control of the ground-truth data and features) rather than on dataset size and model techniques. This argument is largely consistent with the analysis of the studies divided across journal types (psychology vs. engineering), even though the pattern is not well-defined.

<!-- minor improvements to the ending of the paragraph above -->

Third, there is little work on the relevance of the features and how these impact the accuracy of the models. Unfortunately, in many cases, it is impossible to attribute the sources of error to features (as opposed to target emotions obtained from participants or from the modelling architecture, cross-validation, or the size of data), as the studies so far have not compared the feature sets systematically, nor has a comparison of the datasets with identical features has been carried out [@panda2020audio]. In the future, it would be advantageous to systematically assess these sources of error in MER studies to allow us to focus on where and how significant improvements can be made.

<!-- Added at least a reference to Pandas work on survey of features to para above -->

@@

Future reports should contain viable information about the data, models, and success metrics. The modelling process should include a description of cross-validation, feature extraction technique, feature reduction (if used), and actual accuracy measures. We would recommend future studies to use *Matthews correlation coefficient* ($MCC$) for classification studies, as currently there are numerous metrics (*F~1~*, precision, recall, balanced accuracy, etc.), which do not operate in the same range. The situation for regression studies tends to be less volatile, but model accuracy may be reported with error measures (Mean Squared Error, Root Mean Squared Error, or Mean Absolute Error, or $MSE$, $RMSE$, $MAE$), which cannot be converted to relative measure such as $R^2$ without knowing the variance of the data. Again, we would recommend utilising $R^2$ due it being a relative measure and scale invarant to allow comparison across models, datasets, and studies.

<!-- added a recommendation for regression study outputs as well to para above. We could add a mention about R2 adjusted to penalise complex models, but there are other ways to punish for model complexity as well. -->

It would be also highly beneficial to share models and features transparently. If a study relies on published datasets, as most do, sharing the actual model (as code/notebooks) and features would allow for a more direct comparison of model techniques for the same dataset. Also, not all datasets are shared [@akiki2021; @santana2020; @hizlisoy2021mu; @zhang2017fe]. Standardizing reporting of datasets to include subsections explaining detailed information about stimuli (including genre, duration, sampling rate, and encoding format), features (types, extraction/analysis software, quantity of features, transformations, reduction methods), and models (types, tuning parameters, cross-validation) will enable more accurate comparisons between studies. The reporting in @grekow2018au serves as a useful example by providing enough detail in these areas to facilitate reproducibility. Although copyright restrictions may limit the sharing of some datasets, features and examples should be made available through reliable external repositories (e.g., [Zenodo](https://zenodo.org/), [Open Science Framework](https://osf.io/), or [Kaggle datasets](https://osf.io/)).

<!-- Added some examples of where the features/models can be shared -->
